---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    correspondingauthor: true
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1

abstract: |
  We use historical forest harvest data from Alberta, Canada to develop
  a description of the spatial and temporal distributions of the
  availability of forest harvest residues for the period 1990---2015.
  Using a GIS coverage of the road network in Alberta, we developed
  estimates of round-trip haul costs from each harvest location to each of
  three potential bioenergy plant locations. This information was used to
  create supply curves (or cost-availability curves) for each of the three
  potential locations and each of the 26 years. This information is used
  to develop descriptions of the variability in the delivered costs of
  forest residue for bioenergy use. These descriptions of variability are
  historical, but could be useful for potential investors in the bioenergy sector.
keywords:
  - residue
  - road network
  - supply curve
journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: historicalResidueSupply.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-vancouver
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: default
---


```{r setup, include=FALSE, echo = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=FALSE, message=FALSE}
setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis/ms_july")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(scales)
library(gt)
library(quantmod)
library(bookdown)
```


# Introduction

  
Humans have used biofuels since learning to control fire a million or so
years ago. In developed economies, early biofuels (*e.g*. wood,
charcoal, dung, peat) were largely displaced by coal during the
industrial revolution, and by petroleum starting in the mid-nineteenth
century. In recent times, biofuels are once again becoming an important
source of energy because of their potential to reduce greenhouse gas
emissions. In the European Union, the United Kingdom, and Japan, many
heat and power plants have been converted to use wood pellets instead of
coal or natural gas.

Liquid biofuels (*e.g.,* ethanol and biodiesel) have been produced in
commercial quantities from food crops since the 1980s. In some
jurisdictions policies to promote the use of these first-generation
biofuels are being phased out due concerns that they compete with
food production, and that their production may lead to land use change @Campbell2016.

Second generation biofuels rely on processes that convert
lignocellulosic materials (such a corn stover and wood) to liquid
biofuels. An advantage of this process is that the feedstock can be
residue created by food or forest products production, which would otherwise be considered waste. Second
generation biofuel production is still largely in a pilot or
demonstration phase. An interesting example that is planned to use forest
residues is the Silva Green Fuel renewable crude oil plant in Tofte,
Norway which was expected to begin operations in 2021 @Lane2018. As of the
date of writing (`r Sys.Date()`), the plant is not yet operational.

Technologies to pelletize wood are more established than technologies to
produce liquid fuels from wood. It was projected that Canada would
produce 3.8 million Mg of wood pellets in 2021, of which 87% will be
exported, largely to the United Kingdom, Japan, and Europe
@Watters2021.  Wood pellets are replacing coal as the primary energy
source in electricity generating plants, particularly in Europe
@Sterman2018.  

Most of the wood pellet production in Canada comes from
mill residues.  There is interest in expanding the source to the forest
by collecting residue from cutblocks after stemwood harvest: this
practice is well-established in Europe.  Forest
residues are generated as by-products of conventional forest
harvesting operations, and include the tree tops, branches and
non-merchantable logs. These residues are typically left behind in
slash piles, which are burned during winter to mitigate against the
risk of wildfires. Canada has a forestry sector that produces
approximately 20 million oven-dry metric tons of forest residues
annually @Dymond2010.  For the remainder of this paper we will express
biomass using SI units, and all quantities of biomass will be
expressed on an oven-dry basis: 1 oven-dry metric ton is equivalent to
1 Mg.


A number of studies have shown that collection and transportation
costs may be the biggest impediment to widespread use of forest
residues in bioenergy systems
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a bioenergy plant largely depends on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products has a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities
@deAvillez2014 @Spelter2009. This relationship between volatile
harvest volumes and subsequent residue variability is explored in detail by
Niquidet and Friesen @Niquidet2014, who model forest residue
supply in Alberta as a  function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a bioenergy plant's total
costs (@Caputo2005 @Leistritz2007 cited in @Stephen2010),a number of studies have shown that
collection and transportation costs may be the biggest impediment to
widespread use of forest residues in bioenergy systems. reliable feedstock supplies.
Since a
forest-residue based bioenergy plant is dependent on stemwood harvesting
for feedstock, any variability in harvesting directly influences the
volume of residues that would be available. Therefore, consideration of
long-term variability associated with feedstocks in different potential
locations affect the site location decision for a
bioenergy plant.


Past studies regarding bioenergy plant locations have
improved our understanding of feedstock availability and constraints.
One key consideration that has largely been absent from the literature is 
feedstock supply variability over time. Commercial stemwood harvests
vary year-to-year, largely based on market conditions for forest
products. Since a forest-residue based bioenergy plant is dependent on
stemwood harvesting for feedstock, any variability in harvesting
directly influences the volume of residues that would be available.
Therefore, considering the variability associated with feedstocks in
different locations could be important in choosing the best location for a bioenergy plant.

Although few studies consider long-term forest residue variability for
the purposes of locating a potential bioenergy plant,
there are studies that consider variability for other purposes. Some
studies have considered variable feedstock supplies when investigating
whether and when to expand the production of bioenergy from burning
residues @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other studies have explored the context
of variable supplies when making decisions about capital investments in
forest residue-based biorefineries @Cambero2015 @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment decisions
related to investing in a residue preprocessing facility @Chen2012, types of storage infrastructure for bioenergy plant
feedstocks @Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities @Cambero2015.

Our study contributes to the literature by considering variability in
forest residue feedstock supplies for the purpose of selecting the best
location for a bioenergy plant from three potential locations. More
specifically, we use precise information on the location and year of
historical timber harvest in Alberta, Canada to
quantify the spatial and temporal variability of timber harvest, and
thereby relate the harvest information to spatial and temporal
variability of forest residue availability. We develop transportation
cost estimates to all of the historical timber harvest areas based on
cycle time estimates developed for Alberta's road network.
The spatial and temporal variability of forest residues results in
variability in the transport cycle time from the residue collection
areas to the bioenergy plant, which in turn leads to variability in
delivered residue cost. From this information we develop historical cost
curves for collecting and delivering forest residues to each of
the three potential bioenergy plant locations for each of 26 years in
our study period (1990--2015). When considering locations for the
potential bioenergy plants, we conduct sensitivity analysis on 3
different bioenergy plant capacities, as defined by annual forest
residue feedstock requirements: 200 Gg/a corresponding to a small wood
pellet mill, 400 Gg/a corresponding to a large pellet mill or a small
biorefinery, and 800 Gg/a corresponding to a large biorefinery.

The methods used for this study are similar to that used by
FPInnovations for their biomass availability estimates for timber supply
areas in British Columbia (see Ref. @Blackburn2020 for an example). Like us,
FPInnovations develops supply curves, or cost-availability curves, based on
a road network and historical harvests. However, their take on temporal
variability is limited to comparing the averages of two 5-year periods. As well, they
limit their analyses to the boundaries of timber supply areas. The only
boundary that limits our analysis is the provincial border.


Our analysis also shares similarities with work published by Yemshanov
*et al.* @Yemshanov2014. They develop supply curves at national and regional
levels based on a national forest inventory and the location of
cogeneration facilities associated with forest products manufacturing
facilities. However, their analysis is based the reported annual volumes
of processed wood at each manufacturing plant in a single year (2010),
and the corresponding residue delivery cost. Instead, we consider the
spatial location of all the harvest sites over a 26-year period
(1990-2015) and calculate the delivery cost for each harvest site in the
year it was harvested.

In the next section we describe our study area and data sources. We then
describe how we aggregate and analyze the data. We present our
distributional results in the form of marginal delivered biomass cost
for each of the 26 years in the study period, and for each of the three
alternative plant locations. We conclude with a discussion of financial
considerations regarding forest residues as a biofuel feedstock, with
relevance to forest companies, policy makers, and bioenergy investors.



# Study area and data

## Study area

We use the province of Alberta in western Canada as our study area. It
occupies 661 848 km^2^. Its southern boundary is 49°N latitude, northern
boundary is 60°N latitude, eastern boundary is 110°W longitude, and its
western boundary is defined by 120°W longitude and the Great Divide of
the Rocky Mountains.

The southeastern part of the province is largely privately owned
agricultural land, and the northern and western parts are largely
publicly owned forested land. There is, however, a substantial area of
privately owned agricultural land in the Peace River Country of
northwestern Alberta. There are also large areas in National Parks along
the Rocky Mountains and in the far north of the province. The provincial
government identifies the area of mostly publicly owned forested land
(excluding the National Parks) as the Green Area, and the area of mostly
privately owned agricultural land as the White Area (Fig. 1). About 90% of the
harvested timber volume in Alberta comes from the Green Area.

[ Figure 1 about here ]

Our focus for this study is area surrounding the towns of Hinton, Edson,
and Whitecourt in west-central Alberta. These 3 potential bioenergy
plant locations were chosen as they host forest products companies and
are located near forest areas with substantial harvesting activity. We
assume that a potential forest residue-based bioenergy plant would be
located near an existing forest products mill, as timber harvesting is
occurring to supply those mills, and the requisite forest road network
would already be developed. Locating a forest residue processing plant
near a forest products mill would also allow for convenient access to
mill residues.

We used the Weyerhaeuser Company Limited oriented strand board plant in
Edson, the Millar Western Forest Products Ltd. pulp mill in Whitecourt,
and the West Fraser pulp mill in Hinton as the three locations that we
consider for a potential forest residue-based bioenergy plant.

## Data

Several publicly available datasets were to create the township-level
(typically 9.778 km by 9.716 km, or 9 500 ha) summary data used in our
analysis. Most of the data are freely available on the World Wide Web.
All national level data were clipped to the Alberta provincial boundary
for further analysis.

The datasets are introduced here. Their use will be discussed in detail
later.

### Altalis

Altalis manages much of the spatial data created by or for the
Government of Alberta. We use several coverages from their base features
database product @Altalis2021: boundaries for the province, the Green and
White Areas, forest management units (FMUs), forest management agreement
areas (FMAAs), and Alberta Township Survey System (townships provide the
basis for legal land descriptions in Alberta).

### Alberta Geospatial Services

Alberta Geospatial Services is operated by the Alberta Ministry of
Environment and Parks @Alberta2021. We obtained road access coverages from
this database.

### National Forest Information System (NFIS)

The NFIS provides a set of spatial data layers representing forest
properties at a 250 m (6.25 ha) resolution for all of Canada @Beaudoin2017. We
used the merchantable volume, stand age, and species composition
attributes from these data layers.

Another product of NFIS is a set of spatial datasets derived from
satellite data @NFIS2021a. The dataset used for this study shows the
location of timber harvest for each of the years 1985-2015. The dataset
is described in detail by Hermosilla *et al.* @Hermosilla2016. It consists of a raster for
Canada at 30 m (0.09 ha) resolution indicating the year of harvest for
each cell.

The NFIS also provides a web-based individual tree biomass calculator
@NFIS2021b, which we used to calculate stem wood, stem bark, branch, and
foliage biomass for a tree based on province, ecozone, species, diameter
at breast height (DBH) and height.

### Natural Resources Canada

Ref. @Ung2013 describes the tree taper models developed for Canada. We use
these taper models to determine merchantable stemwood volumes and the
amount of forest residues (branches and tops) available from harvested
areas in each of the townships in each year.

### National Forestry Database (NFD)

The NFD @CCFM2021 provided the timber harvest volumes and total harvest
areas from Alberta for the years 1990-2015. We used this information to
adjust the spatial NFIS volumes mentioned above to better reflect the
actual aggregated harvest volumes observed in Alberta.

### Forest management plans

The Alberta government requires Forest Management Agreement (FMA) holders to develop a forest
management plan for the FMAA. The forest management plans were used to
obtain information on utilization standards and average piece size of
harvested trees for the FMAA @Alberta2021.

### Agriculture and Agri-Food Canada

The boundaries of Canada's terrestrial ecozones can be found on
Agriculture and Agri-Food Canada's website @Canada2017. Ecozone is an input
into the NFIS individual tree biomass calculator @NFIS2021b, described in
section 2.2.3.


# Data aggregation and modeling 


We quantify the spatial and temporal variability of the costs associated
with collecting and transporting the forest residues left following
stemwood harvest. The variability of costs is largely related to variability in the transport distance from the residue collection areas to the
bioenergy plant.

Most of the effort expended on this study was spent in aggregating
township-level summaries of data for use by the model. Much of this data
processing was the manipulation of spatial data, which was done using
the QGIS geographic information system (GIS) software @QGISorg2021 and
add-ons. Data summaries and statistical analysis were done using R
@R2021 with RStudio @RStudioTeam2021 and the tidyverse @RStudioTeam2021 
collection of R packages.

We develop a model based on the assumption that forest residue
extractors follow a greedy optimization procedure: they take the
cheapest (closest) residue first, and stop when they have enough to
satisfy the annual biomass requirements for each of the three scales of
bioenergy plants: 200 Gg/a, 400 Gg/a, and 800 Gg/a. Township level
residue yield (Mg·ha^-1^), cycle times (h), and harvest areas (ha) are
the three essential summary parameters used by the greedy optimization
model.


## Data aggregation

### Cycle time estimates


The provincial road network was divided into four classes: primary
highways, other paved roads, two-lane gravel roads, and one-lane gravel
or dirt roads. Loaded and unloaded travel speeds were assumed for each
road class, according to Table 1. These travel speeds are based on those
assumed by FPInnovations for the Dawson Creek Timber Supply Area in the
neighbouring province of British Columbia @Blackburn2020.

The harmonic mean of loaded and unloaded speeds was calculated using Eq.
1.

$$\bar{s_{h}} = \left( \frac{s_{l}^{- 1} + s_{u}^{- 1}}{2} \right)^{- 1}$$
Where $s_{l}$ is the loaded speed and $s_{u}$ is the unloaded speed.
Use of harmonic mean allows us to calculate the correct cycle time: the
time required to travel from the mill to the collection site at the
cutblock, and back to the mill. This provides the basis for our
estimates of haul cost.

Figure 2 illustrates the road network in Alberta and estimates of cycle
time based on the average of loaded and unloaded speeds from Table 1.
The road network in the White Area of the province is very dense and
largely arranged as a grid defined by the Alberta Township Survey
System. Most of the roads in the White Area are public. The road network
in the Green Area is less dense and many of the roads are built and
maintained by resource industries (*e.g.,* forestry, petroleum, and
natural gas). The pattern of the road network is not as regular as in
the White Area.

[ Figure 2 about here ]

One-way haul times based on the harmonic mean travel speed were
generated using the QGIS Network Analysis Toolbox 3 plug-in (QNEAT3)
@Raffler2018, specifically its "iso-area as interpolation (from point)"
algorithm. In our case, we used the algorithm to determine the minimum
one-way travel time (using the harmonic mean of loaded and unloaded
speeds) along the road network from each of our centers to each cell of
a provincial level raster at a 200 m resolution. Off-road travel was
assumed to be at 5 km·h^-1^. The Dijkstra algorithm @Dijkstra1959 is the basis
for travel time minimization in QNEAT3. Cycle time is double the one-way
haul time with an extra hour added for loading and unloading.

QGIS zonal statistics were used to calculate the median cycle time to
each township, from each of the three potential bioenergy plants. The
cycle time in Figure 2 was limited to 14 hours because, in Alberta,
drivers of commercial vehicles are not permitted to drive more than 13
hours in a 24-hour period and can spend no more than 14 hours on-duty,
including loading and unloading times.

### Biomass collection and transportation costs

All costs and prices used in this study are in Canadian dollars (\$). At
the time of analysis (2021-11-05), the exchange rate between Canadian
(CAD) and US dollars (USD) was 0.8031 USD/CAD.

We assume that the biomass available for collection at a cutblock
includes the tops, branches, and foliage left at a roadside landing
after processing by a stroke delimber. The delimber is assumed to cut
the top of the tree at the diameter specified in the utilization
standard specific to the harvest area. Following Ref. @Roser2013, the cost
of collecting residues left behind by a roadside stroke delimber, which
includes pre-piling, chipping, road maintenance, supervision, and
loading (but does not consider transport) is 41.60 \$·Mg^-1^.

We assume that the residue will be transported from the forest using a
live floor chip van with 100 m^3^ capacity. The USDA Forest Service's
Forest Residue Transportation Costing Model @Rummer2005 assumes a default
solid volume factor of 0.3 for biomass: a 100 m^3^ capacity
trailer could hold the equivalent of 30 m^3^ of solid biomass. This
equates to 11.40 Mg of biomass using the specific gravity of lodgepole
pine wood and bark on a green volume basis (0.38) (Ref. @Miles2009).

Timber Tracks produces a publication describing forestry equipment
hourly rates @TT2020. The rate reported for a tandem tractor with a
tandem trailer was 157.31 \$·h^-1^. We assume that the tractor-trailer
combination used to haul biomass would cost a similar amount.

Therefore, the cost of transporting residual biomass was set to 13.80
\$·Mg^-1^·h^-1^.

### Aggregation of harvested area

The areas harvested by township in 2009 according to Refs. @NFIS2021a @Hermosilla2016
are shown in relationship to the three potential bioenergy plant
locations and the Green Area in Figure 1. Similar aggregated data were
created for each of the years 1990--2015. For each township, and each of
the years between 1990 and 2015, we store the area harvested (ha) for
use in our analysis. We can see from this map that a great deal of
timber harvest took place near our potential bioenergy plant locations
in 2009.

These data were created by using QGIS zonal statistics to sum the area
harvested in each township in each year based on the 30 m resolution
harvest data @NFIS2021a.

### Conversion of harvested area to harvest volume and available biomass

Much detailed work was done to reflect the variability in utilization
standards between the FMUs, determination of characteristics for the
average broadleaf and the average needleleaf tree in each township, and
the spatial and temporal distribution of stemwood harvests. The
description of the procedures used for this are in the Appendix.



## Aggregation of summary data files

The information collected above was collated into a data set which had
a record for each combination of township, plant location, year, and
capacity. The fields in the data set were of township, plant location,
year, and capacity, cycle time, harvest area, and residue yield. This
summary data set is the input to the greedy optimization model.

## Modeling

Our model assumes that residue extraction is opportunistic. The
historical data showing when and where forest managers have harvested
stemwood and our model assumes that residue extraction would occur
afterwards, in the same year as timber harvest. The forest manager's
decision to harvest stemwood is made independently of the biomass value
of the forest residue, and our cost estimates do not include a payment
to the forest manager for the residue. This assumption is appropriate,
given that biomass extractors provide a service to the forest managers,
who would otherwise have to incur the costs of piling and burning the
residue if it was not extracted.

Every township is associated with a cycle time required to transport
forest residue from the township to each of the three potential
bioenergy plant locations, for each of the 26 years in the study period.
For each plant location and year, the model examines all the townships
in the province that have a positive residual volume and begins the
creation of cumulative residual biomass curves starting with the lowest
cycle time first. These lists of cycle time and cumulative biomass
provide the data for the marginal cycle time curves, which can then be
converted to marginal cost curves using the hourly rate and payload for
forest residue transport trucks.

The model uses a greedy algorithm whereby the lowest cost residue for a
location is extracted first. The model is summarized using pseudocode in
Figure 3.

[ Figure 3 about here ]



# Results

## Distributional results

Our results assume that only one of the three potential bioenergy plants
would be built. If more than one plant was to be built, the residue
collection areas would overlap, thereby reducing the available residue
for the both plants. Figures 4 and 5 display marginal cycle time (h)
and marginal costs (\$·Mg^-1^) for each of the 26 years in the study
period, and for the 3 potential plant locations. Curves for 1990, 2009,
and 2005 are highlighted represented the years of minimum, median, and
maximum provincial harvest volume in the period. The inter-year
variation in marginal costs increases as the cumulative amount of
residue collected increases. We can also see that the relative position
of the years in terms of cost varies between the plant locations:
*e.g.,* the year 2005 was the year of highest provincial harvest, but
from a residue availability standpoint was not as good for Hinton as it
was for Edson and Whitecourt.

[ Figure 4 about here ]

[ Figure 5 about here ]

Figure 6 is presents another view of theinterannual variation in the
distribution of costs between the plant locations and assumed capacity:
the boxplots display the quartiles, and the minimum and maximum costs
require to supply a bioenergy plant of a given capacity at each of the
locations in each of the years. Here again, we can see the variability
increasing as plant capacity increases.

[ Figure 6 about here ]


Figure 7 shows the average cost (\$·Mg^-1^) for each potential location,
plant capacity, and year. Overall, the average costs in Edson and
Whitecourt are quite similar. Hinton is usually more expensive across
years and capacities. The differences become clearer at a capacity of
800 Gg/a, for which Whitecourt is the lowest cost location for most
years.

[ Figure 7 about here ]

Table 2 presents summary statistics over the 26 years for the locations
and capacities. Average costs and the interannual variability of costs
increase with increasing capacity. The variation in costs for Hinton is
greater than for Whitecourt and Edson, particularly for the 800 Gg/a
capacity.

[ Table 2 about here ]

Figure 8 shows the areas that would have been needed to supply a
bioenergy plant of varying capacities in Whitecourt in the year of
median harvest (2009) in relation to the proportional township area
harvested. Note that there is some residue being collected from the
White Area, but most is being collected from the Green Area.

[ Figure 8 about here ]


Figure 9 compares the supply areas for 2005, 2009, and 1990 (high,
medium, and low harvest years, respectively) for Whitecourt. The area
required to supply the bioenergy plant increases greatly in low timber
harvest years, particularly for the 800 Gg/a capacity. Similar maps
have been produced for the other locations and all 26 years and are
available at the University of Alberta's Education and Research Archive
(ERA). **TO BE DONE** Notice that the supply areas include areas outside the
Green Area with very little harvest. The contours indicate the cycle
time associated with each of the capacities. Because of the cost of
moving equipment such as the grinder and loader, it is unlikely that the
small, dispersed harvested areas would be accessed, but the volumes are
small enough that we ignore them. We have created animations of these
contour maps for all 26 years in the study period for the 3 locations
(available at ERA). These provide another way of visualizing the
interannual variation.

[ Figure 9 about here ]

Whitecourt is lowest cost location for most years at all capacities.
There is a noticeable spike in costs 1999, which may be related to
salvage harvest following the 1998 Virginia Hills wildfire @Prepas2001. Note
that this spike also affects Hinton and Edson at 800 Gg/a capacity,
because they reach into the Whitecourt area at that capacity level. The average cost of
delivered forest residues are lower for the lower capacity plants, as
they do not need to collect residues from as far to meet their biomass
input requirements. In other words, the average cycle time is lower for
the lower capacity bioenergy plants, which results in a lower average
delivered residue cost.

Overall, the results show that there is a high level of interannual
variability in the costs of retrieving a fixed amount of biomass from
timber harvesting residues. The variability is related to the area and
location of harvest, which varies from year-to-year, and directly
impacts transportation costs.



# Conclusions

Our model demonstrates the spatial and temporal variation in the
availability of forest residue for bioenergy production. We did this
using the location of timber harvest areas in the Canadian province of
Alberta for the years 1990--2015. We conducted this study to answer
three questions critical to bioenergy investors and policy makers. We
now discuss our findings for each of these questions in turn.

**Which areas are best for locating a potential bioenergy plant when
considering residue collection and transport costs?** We focus on the
area surrounding the towns of Hinton, Edson, and Whitecourt in
west-central Alberta. These three potential bioenergy plant locations
were chosen as they host several forest products companies and are
located near forest areas with substantial harvesting activity. For each
location, we conduct sensitivity analysis on three alternative bioenergy
plant capacities as measured by residue feedstock requirements: 200
Gg/a, 400 Gg/a, and 800 Gg/a. For the 200 Gg/a
capacity the three locations are almost indistinguishable in terms of
average delivered residue cost (\~ 87 \$·Mg^-1^). At 400 Gg/a,
Edson is the low-cost location (98.6 \$·Mg^-1^). At 800 Gg/a,
Whitecourt is clearly the low-cost location (122 \$·Mg^-1^), however, it
is important to note that the average cost of supplying an 800 Gg/a
plant is over 40% greater than supplying a 200 Gg/a plant.

**How much residue volume is available over what area?** The residue
supply areas for the Whitecourt location shown in Figure 9 suggest that
if a 400 or 800 Gg/a bioenergy plant was to be constructed, there
would only be enough available residue to construct one plant in our
study area. Whereas if a 200 Gg/a plant was deemed more viable, our
findings suggest that perhaps two plants could be constructed in the study area
-- one in Whitecourt and one in Hinton. This makes for an interesting
investment choice, since a single 800 Gg/a bioenergy plant would
likely face an average delivered residue cost that is over 40% higher
than the cost for the 200 Gg/a plants. Recall from above that 200
Gg/a is the appropriate scale for a pellet plant and 800 Gg/a
is more representative of a biorefinery. Given that dried and compacted
pellets are cheaper to transport than moist and bulky residues, it could
be more viable to first transport residues to pellet plants, and then
transport the pellets (perhaps via rail) from multiple pellet plants to
a biorefinery. The optimal result of this analysis will depend on the
cost of pellet production relative to the cost savings (transportation
and production) to the biorefinery from using pellets instead of
residues. .

**Would residue volumes be consistently available on a year-to-year
basis?**  Given that a forest residue-based bioenergy plant is dependent
on forest harvesting, transportation cost increases in years when
harvesting levels are low or harvest areas are further from the
bioenergy plant. Our results showed considerable variability in both
marginal and average residue extraction costs from year-to-year for the
400 and 800 Gg/a capacities, suggesting that residue cost
variability is an important source of risk for medium and large
bioenergy plants. Our finding that the cost variability is lower for the
200 Gg/a might once again suggest that a 200 Gg/a is a more
appropriate scale for a forest residue-based bioenergy plant in Alberta.

Our conclusions are based on an analysis of historical harvest levels,
which might not be indicative of the future. However, we feel confident
that our findings will be relevant into the future largely because
public forest in Alberta is managed according to the sustained yield
paradigm, which involves the determination of stable allowable annual
cuts over a 200-year planning period. For evidence of this stability,
Figure 2 shows that harvest volume rises until the year 1994, which
coincides with the last of the major forest product mills in Alberta
being constructed in the early-1990s, and after this date harvest levels
flatten out and become fairly stable. Transportation costs can also be
expected to be fairly stable in the future because long-term forest
management plans in Alberta require a balanced log haul distance. In
other words, the Alberta government would not approve a plan whereby
forestry firms concentrate harvesting operations such that haul
distances were to significantly increase in the future. Hence, it is
reasonable to assume that in the future, both average annual harvest
levels and average annual transport distances -- and thereby average
annual forest residue costs -- will be close to the historical values
from our analysis.

Although we are confident that our conclusions will hold in the future
over the long-term, it is nonetheless possible that over the long
operating life of a forest residue-based bioenergy plant, there could be
significant events that could impact residue availability over the
short-term -- *e.g.*, bad weather which prevents forest harvest and/or
residue extraction, temporary mill shutdowns, and forest fires. This
possibility makes it important for a forest residue-based bioenergy
plant to have a comprehensive feedstock risk management program. If the
bioenergy conversion technology allows it, this risk management program
could consider creating a portfolio of different sources of feedstock.
Such a portfolio approach might consider locating the bioenergy plant
where it could also access other types of feedstock -- such as
agricultural residues and/or purpose-grown feedstocks -- during periods
when availability of forest residues is low. This approach is sometimes
called "derisking" in the feedstock supply literature, whereby a supply
of "buffer" feedstock is contracted to keep the bioenergy plant running
during years when normal residue supplies are low @Golecha2016. Ref. @Zheng2021
discusses the amount and variability of production of agricultural
residue in Alberta; and Ref. @Shooshtarian2018 discusses the
financial viability of establishing fast-growing hybrid poplar
plantations in Alberta for use as a purpose-grown bioenergy feedstock.

When considering portfolios of different bioenergy feedstocks, an
important factor will be the average delivered cost for each feedstock.
Our estimates for average delivered forest residue costs range from
approximately 87 \$·Mg^-1^ for the 200 Gg/a capacity to
approximately 122 \$·Mg^-1^ for the 800 Gg/a capacity. Previous
studies have estimated the cost of agricultural residue (straw) to fall
within this range, with a plant gate cost in Alberta of 95.33 \$·Mg^-1^
for a plant capacity of 150 Gg/a @Sultana2010. Given how
close the cost of agricultural residues is to the cost of forest
residues, these two feedstocks could be a good fit for a potential
feedstock portfolio, assuming the bioenergy plant could use both types
of feedstock. However, a previous study estimated that purpose-grown
hybrid poplar in Alberta has a much higher delivered cost of 202
\$·Mg^-1^ (125 \$·Mg^-1^ to grow the trees, plus 77 \$·Mg^-1^ to
harvest, grind, and transport the trees to the bioenergy plant)
@Shooshtarian2018, suggesting that hybrid poplar would have a
more limited role (if used at all) within a feedstock portfolio. In the
future we hope to build upon this research by optimizing the allocation
of different feedstocks within a bioenergy feedstock portfolio.

Although forest residues are utilized for bioenergy in other parts of
the world, they are still largely considered a waste product in Canada.
Given that feedstock costs are critical to the economic viability of any
bioenergy project, in order for forest residues to be extracted in
Canada we will likely need to see improvements in bioenergy production
technology and/or increases in future biofuel prices. Governments can
play a role in improving bioenergy production technology through
investments research and development, and can impact future biofuel
prices through public policy.

On the Canadian public policy front, governments have set renewable fuel
mandates and subsidize production of bioethanol and biodiesel, thereby
impacting future biofuel prices @Campbell2016. The
continuation of such subsidies will likely be an important factor for
the emergence of a second-generation or advancedbiofuel sector in
Canada. Another factor that could provide important incentives for
advanced biofuels is the Government of Canada\'s Clean Fuel Standard,
which is proposed to come into force in 2022, and which will require
transportation fuel suppliers to lower the carbon intensity of their
fuels @Canada2020. Although these policies are
focused more on liquid biofuels, there are also new policies which could
promote burning pellets. Indeed, the Government of Canada recently
implemented carbon pricing regulations @CanLII2018 which could
provide incentives to use forest residue pellets in heat and power
plants instead of coal and natural gas. In this case, since the pellets
would likely generate lower greenhouse gas emissions than the fossil
fuels they displace, the heat and power produced would thereby incur
lower carbon taxes, thus improving the financial returns to using
pellets.


# Data availability

The data compiled to the township level are currently available on
github(<https://github.com/gwa-uab/histressup/tree/main/data>). Before
publication it will be placed on ERA with a DOI.

# Acknowledgements

This research was supported by funding from the Canada First Research
Excellence Fund as part of the University of Alberta's Future Energy
Systems research initiative (Project \# CFREF-2015-00001).

Dennis Gray contributed to data compilation.

This work started as an extension to Irene Onyango's Master of
Agriculture research.

Thanks to Marty Luckert, Feng
Qiu, Grant Hauer, and Wenbei Zhang for helpful discussion.




References {#references .unnumbered}
==========
