---
title: Spatial and temporal variability in economic availability of residual biomass from timber harvest in west-central Alberta, Canada (1990--2015)
subtitle: Appendix
author:
  - name: Glen W. Armstrong
    email: gwa@ualberta.ca
    affiliation: rr
    correspondingauthor: true
  - name: Jay A. Anderson
    email: jaya@ualberta.ca
    affiliation: rees
  - name: Alex McPhee
    email: awmcphee@ualberta.ca
    affiliation: rees
address:
  - code: rr
    address: Department of Renewable Resources, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1
  - code: rees
    address: Department of Resource Economics and Environmental Sociology, University of Alberta, Edmonton, Alberta, Canada, T6G 2H1

journal: "Biomass & Bioenergy"
date: "`r Sys.Date()`"
classoption: preprint, 3p, authoryear
bibliography: historicalResidueSupply.bib
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/elsevier-vancouver
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: default
---

Please make sure that your manuscript follows the guidelines in the 
Guide for Authors of the relevant journal. It is not necessary to 
typeset your manuscript in exactly the same way as an article, 
unless you are submitting to a camera-ready copy (CRC) journal.

For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>

# Bibliography styles

Here are two sample references: @Feynman1963118 [@Dirac1953888].

By default, natbib will be used with the `authoryear` style, set in `classoption` variable in YAML. 
You can sets extra options with `natbiboptions` variable in YAML header. Example 
```yaml
natbiboptions: longnamesfirst,angle,semicolon
```

There are various more specific bibliography styles available at
<https://support.stmdocs.in/wiki/index.php?title=Model-wise_bibliographic_style_files>. 
To use one of these, add it in the header using, for example, `biblio-style: model1-num-names`.

## Using CSL 

If `citation_package` is set to `default` in `elsevier_article()`, then pandoc is used for citations instead of `natbib`. In this case, the `csl` option is used to format the references. Alternative `csl` files are available from <https://www.zotero.org/styles?q=elsevier>. These can be downloaded
and stored locally, or the url can be used as in the example header.

# Equations

Here is an equation:
$$ 
  f_{X}(x) = \left(\frac{\alpha}{\beta}\right)
  \left(\frac{x}{\beta}\right)^{\alpha-1}
  e^{-\left(\frac{x}{\beta}\right)^{\alpha}}; 
  \alpha,\beta,x > 0 .
$$

Here is another:
\begin{align}
  a^2+b^2=c^2.
\end{align}

Inline equations: $\sum_{i = 2}^\infty\{\alpha_i^\beta\}$

# Figures and tables

Figure \ref{fig2} is generated using an R chunk.

```{r fig2, fig.width = 5, fig.height = 5, fig.align='center', out.width="50%", fig.cap = "\\label{fig2}A meaningless scatterplot.", echo = FALSE}
plot(runif(25), runif(25))
```

# Tables coming from R

Tables can also be generated using R chunks, as shown in Table \ref{tab1} for example.

```{r tab1, echo = TRUE}
knitr::kable(head(mtcars)[,1:4], 
    caption = "\\label{tab1}Caption centered above table"
)
```

# References {-}

